Mergeable Types

### 2017-02-28 paper/examples.tex
fix wrapfig
minor edits: spelling fixes

A rope is a data structure that faciliates efficient implementations
of operations such as concatenation, splitting, lookup and insertion
at a random position etc., for a list of values in a distributed
setting. The values are usually characters, with their lists being
strings. The data structure is a binary tree, whose leaf nodes store
substrings of a larger string such that their left-to-right (in-order)
concatenation yeilds the larger string. The idea is to facilitate
efficient lookups and insertions at random positions in the string by
storing position indexes in the internal nodes of the tree.

@@

One feature of the mergeable rope implementation that is characterisic
of mergeable types in general is compositionality; a rope is a
polymorphic mergeable data type that composes mergeability over its
polymorphic constituents.   Compositionality is concretized by the
\C{MRope} signature , which requires the type of items in the list
(\C{A.t}) to also be mergeable (\C{A.merge}).

@@

A mergeable list of integers can be composed using mergeable lists and
mergeable integers.  Concretely, it is a composition of \C{MList} and
\C{Int} types (i.e., \C{MList(Int)}), where \C{Int.merge} is the same
as \C{Counter.merge} (Sec.~\ref{sec:intro}).  An alternative
instantiation of \C{MList} is for a quantity type
(\C{Qty}) that supports \C{add} and \C{subtract} operations, but
\begin{wrapfigure}{l}{.45\textwidth}
  \centering
  \includegraphics[scale=0.4]{Figures/list-eg}
  \caption{Lists of mergeable values are mergeable. }
  \label{fig:list-eg}
  \vspace*{-.2in}
\end{wrapfigure}
defines its \C{merge} opeation a little differently:
\begin{ocaml}
let merge lca v1 v2 = if v1 = lca then v2
    else (if v2 = lca then v1 else max v1 v2)
\end{ocaml}
\C{MList(Qty)}, like \C{MList(Int)}, is a list of integers, and has
the same semantics under a sequential execution. Their differences
however manifest under a concurrent execution, where both adopt
different methods to reconcile concurrent updates.

@@

A non-trivial demonstration illustrating the compositional power of
mergeable types is shopping list application, that allows its users to
collaboratively build a shopping list (listing items in the decreasing
order of their priority), by adding new items with a
specified quantity, deleting existing items, and increasing or
\begin{wrapfigure}{l}{.5\textwidth}
  \centering
  \includegraphics[scale=0.3]{Figures/shopping-list}
  \caption{Merging concurrent shopping lists}
  \label{fig:shopping-list}
  \vspace*{-.2in}
\end{wrapfigure}
decreasing their quantity. An item (\C{Item.t}) is represented as a
tuple of its name (a string) and the quantity (A \C{Qty.t}). The merge
function for \C{Item.t} merges the quantities of items (using
\C{Qty.merge}) having the same names. Items with different names are
considered distinct. A shopping list is a mergeable list of items,
i.e., \C{MList(Item)} (a grocery shopping list may not be large enough
to justify using ropes, but \C{MRope(MList(Item))} is definitely a
possibility).  \C{MList(Item).merge} automatically lifts the merge
semantics for items to a merge semantics for lists of items.
Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
users, Alice and Bob, concurrently edit a shopping list whose initial
contents are \C{milk} and \C{eggs}. While both simultaneously update
the quantity of eggs, Bob also removes \C{milk} and inserts \C{candy}.
Fig.~\ref{fig:shopping-list-code} shows the \name code for two
concurrent sessions. The merged shopping list contains \C{eggs} and
\C{candy}, where the quantity of eggs is obtained by merging Alice's
and Bob's quantity as per the definition of \C{Qty.merge}.

>>>>

A rope is a data structure that facilitates efficient implementations
of operations such as concatenation, splitting, lookup and insertion
at a random position etc., for a list of values in a distributed
setting. The values are usually characters, with their lists being
strings. The data structure is a binary tree, whose leaf nodes store
substrings of a larger string such that their left-to-right (in-order)
concatenation yeilds the larger string. The idea is to facilitate
efficient lookups and insertions at random positions in the string by
storing position indexes in the internal nodes of the tree.

@@

One feature of the mergeable rope implementation that is characteristic
of mergeable types in general is compositionality; a rope is a
polymorphic mergeable data type that composes mergeability over its
polymorphic constituents.   Compositionality is concretized by the
\C{MRope} signature , which requires the type of items in the list
(\C{A.t}) to also be mergeable (\C{A.merge}).

@@

A mergeable list of integers can be composed using mergeable lists and
mergeable integers.  Concretely, it is a composition of \C{MList} and
\C{Int} types (i.e., \C{MList(Int)}), where \C{Int.merge} is the same
as \C{Counter.merge} (Sec.~\ref{sec:intro}).  An alternative
instantiation of \C{MList} is for a quantity type
(\C{Qty}) that supports \C{add} and \C{subtract} operations, but
\begin{wrapfigure}{L}{.45\textwidth}
  \centering
  \includegraphics[scale=0.4]{Figures/list-eg}
  \caption{Lists of mergeable values are mergeable. }
  \label{fig:list-eg}
  \vspace*{-.2in}
\end{wrapfigure}
defines its \C{merge} operation a little differently:
\begin{ocaml}
let merge lca v1 v2 = if v1 = lca then v2
    else (if v2 = lca then v1 else max v1 v2)
\end{ocaml}
\C{MList(Qty)}, like \C{MList(Int)}, is a list of integers, and has
the same semantics under a sequential execution. Their differences
however manifest under a concurrent execution, where both adopt
different methods to reconcile concurrent updates.

@@

A non-trivial demonstration illustrating the compositional power of
mergeable types is shopping list application, that allows its users to
collaboratively build a shopping list (listing items in the decreasing
order of their priority), by adding new items with a
specified quantity, deleting existing items, and increasing or
\begin{wrapfigure}{L}{.5\textwidth}
  \centering
  \includegraphics[scale=0.3]{Figures/shopping-list}
  \caption{Merging concurrent shopping lists}
  \label{fig:shopping-list}
  \vspace*{-.2in}
\end{wrapfigure}
decreasing their quantity. An item (\C{Item.t}) is represented as a
tuple of its name (a string) and the quantity (A \C{Qty.t}). The merge
function for \C{Item.t} merges the quantities of items (using
\C{Qty.merge}) having the same names. Items with different names are
considered distinct. A shopping list is a mergeable list of items,
i.e., \C{MList(Item)} (a grocery shopping list may not be large enough
to justify using ropes, but \C{MRope(MList(Item))} is definitely a
possibility).  \C{MList(Item).merge} automatically lifts the merge
semantics for items to a merge semantics for lists of items.
Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
users, Alice and Bob, concurrently edit a shopping list whose initial
contents are \C{milk} and \C{eggs}. While both simultaneously update
the quantity of eggs, Bob also removes \C{milk} and inserts \C{candy}.
Fig.~\ref{fig:shopping-list-code} shows the \name code for two
concurrent sessions. The merged shopping list contains \C{eggs} and
\C{candy}, where the quantity of eggs is obtained by merging Alice's
and Bob's quantity as per the definition of \C{Qty.merge}.
